---
title: "Introduction to exvatools"
author: Enrique Feás
date: December, 2022
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to exvatools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

International input-output tables are useful statistical tools: they show us
how different sectors in  different countries interact with each other. This
is particularly relevant when analyzing international trade, given that half of
today's trade is composed of intermediate goods and gross flows become less and
less representative of reality. Moreover, international input-output tables 
show the increasing relations between services and goods.

The problem is that handling big input-output tables is complicated. This is
where **exvatools** comes to the rescue. Its purpose is double:

* To facilitate the extraction and processing of the most relevant 
  international input tables available today, through easy commands.
* To provide tools to decompose the value added in exports
  according to different methodologies in the literature.

## International input-output analysis tools

### Getting raw data

There are a few international input-output tables available. Two of the most
renowned in economic literature are:

* The OECD's Trade in Value Added database (TiVA), based on the Inter-Country
  Input-Output Tables (ICIO)
* The University of Groningen's World Input Output Database (WIOD), based on
  the World Input Output Tables.

The advantage of these databases is twofold: they are quite rich in terms of
countries and sectors, and they are directly downloadable from the web page of
their supporting institutions.

Source data is normally provided in zipped files containing comma-delimited 
files (`.csv`), Excel files (`.xlsx`) or R data files (`.rData`), which
**exvatools** can handle with the support of other CRAN packages.

In order to use **exvatools** with official data, we must first download the
source files to our computer. For instance, if we want to use the last 
edition of the OECD (2021) with data up to 2018, we should download the
file "ICIO_2015-2018.zip" (96 MB) from the
[ICIO web page](https://www.oecd.org/sti/ind/inter-country-input-output-tables.htm) 
containing the required `.csv` files.

From here, **exvatools** will take care of the rest: it will extract the raw 
data from the zip file and create the basic input-output matrices, embedded
in objects that will be the basis for further analysis.

### Installing exvatools

Installing **exvatools** from the CRAN repository follows the usual
procedure:

```{r install, eval=FALSE, echo=TRUE}
install.packages("exvatools")
```

To make **exvatools** available:
```{r library}
library(exvatools)
```

### Importing data and creating basic input-output matrices

Once we have downloaded the raw data, we can obtain the basic input-output
matrices in an object `wio` (for 'world input-output') with the command 
`make_wio()`. We just need the input-output edition we want to use (the one
we have downloaded), the year and the folder where we saved our zip file 
(just the directory). For instance, if we had downloaded it in
`C:\Users\Username\Documents\R`, we would just need to type the following
command (mind the direction of the slashes in **R**):

```{r make_wio, eval=FALSE, echo=TRUE}
wio <- make_wio("icio2021", year = 2018, 
                src_dir = "C:/Users/Username/Documents/R")
```

As the dimension of these input-output matrices is big, for didactic purposes
we will use a test input-output database, with non-real data. **exvatools**
includes two test tables, `"iciotest"` (which follows a ICIO-type structure)
and `"wiodtest"` (which follows a WIOD-type structure).
```{r make_wio_test}
wio <- make_wio("iciotest")
```

We can easily check the contents of the created `wio` object with `summary()`:
```{r summary_wio}
summary(wio)
```

### Operating with input-output matrices

**exvatools** provides multiple commands that make operating with international
input-output tables extremely easy: thus, we can multiply a diagonal matrix by
an ordinary one with `dmult()`, an ordinary by a diagonal with `multd()`, or 
make a block-by-block Hadamard product of matrices with `hmult()`.

We can also easily obtain a block diagonal matrix with `bkd()`, a block 
off-diagonal matrix with `bkoffd()`, or a diagonal matrix with the sums of all
columns with `diagcs()`, 

Additionally, as we are always operating with named rows and columns with
names of countries and sectors, we have improved commands to consolidate
matrices and provide names, like `rsums()` to sum rows, `csums()` to sum 
columns, `sumnrow()` to sum every nth row of a matrix, `sumncol()` to sum every
nth column, `sumgrows()` to sum groups of rows of a particular size,
`sumncols()` to do the same with columns, etc.

For instance, let us check that the production `X` is equivalent to the product
of the global Leontief inverse matrix `B` and the final demand `Y`:
```{r BY}
BY <- wio$B %*% wio$Y
```

We can sum the rows and check that it coincides with the production vector:
```{r sum_WBY}
BY <- rsums(BY, "BY")
print(cbind(head(BY, 10), head(wio$X, 10)))
```

The OECD ICIO tables have a particular feature: two big industrial countries,
China and Mexico, are disaggregated in two for calculation purposes. This
requires a tool to consolidate data once all calculations are done. The 
command `meld()` takes care of that.

For instance, to calculate the value added absorbed abroad we need to 
multiply the value added coefficients matrix `V`-hat (represented here
with `W`) by the global inverse matrix `B` by the final demand matrix `Y`, 
and then exclude the value added absorbed domestically. This can be easily done
with a few commands.
```{r WBY}
# To calculate all value added induced by demand:
VBY <- dmult(wio$W, wio$B) %*% wio$Y
VBY
```

We can see that rows for Mexico and China are disaggregated. We can now
meld them:

```{r meld}
VBY <- meld(VBY)
VBY
```

We just want the value added absorbed abroad (sometimes referred to as 
value added exported). For that we need the block off-diagonal matrix of
`VBY`, that we can produce with `bkoffd()`:

```{r bkoffd}
bkoffd(VBY)
```

## Value added in exports

The model we have used allows us to express value added induced by demand. This
is the idea behind the calculation of value added in exports, i.e., induced by
the demand of exports. In this case, we need not only to obtain the value 
added, but to estimate the effect of the multiple times that intermediate 
products cross the international borders.

This is why several methods have appeared to calculate full decomposition of
value added in exports, distinguishing what part of value added is *pure*
value added and which one is double counting, and also, within the *pure*
double counting, what part is really exported and what part eventually
returns back to the exporting country ('false' exports).

### Alternative decompositions of value added in exports

There are several methodologies in the economic literature, and **exvatools**
includes the most complete ones: Koopman et al. (2014), Wang et al. (2013),
Borin and Mancini (2019) and Miroudot and Ye (2021).

For instance, to  create create a full decomposition of value added 
in exports of Spain using the method of Borin and Mancini (2019), 
using a source-based approach, we would type::

```{r make_exvadec ESP}
exvabm <- make_exvadec(wio, exporter = "ESP", method = "bm_src")
```

The advantage is that, once we have obtained a decomposition, we can play 
with the results in terms of sectors and countries of destination just
using the command `get_exvadec_bkdown()`. For instance, to select the
value added in Spanish exports of services (including construction) 
to the United States, we just have to type: 

```{r bkdown_esp_srvwc_usa}
get_exvadec_bkdown(exvabm, exporter = "ESP", 
                   sector = "SRVWC", importer = "USA")
```

An alternative (although methodologically inferior) decomposition would be
the Wang et al. (2013) decomposition. In this case, instead of the normal 
decomposition, we will use the `"terms"` output. (that shows the 16 terms that
compose the value added in exports):

```{r make_exvadec_wwz}
exvawwz <- make_exvadec(wio, exporter = "all", method = "wwz",
                        output = "terms", quiet = TRUE)
```

Note that here we have selected `export = all`, that produces decompositions 
for all countries (not only a specific one), and we have also used the option 
`quiet = TRUE`, that produces a silent output.

We can check any exporting country, any sector, and any destination country. For
instance, we can produce the decomposition of the value added in US exports to
China for the manufacturing sector:

```{r bkdown_usa_manuf_chn}
get_exvadec_bkdown(exvawwz, exporter = "USA", 
                sector = "MANUF", importer = "CHN")
```

### Direction of value added in exports: origin and destination

We have seen that the foreign content in Spanish exports amounts to USD 2501.78
million. Where does it come from? If we can pass without the detailed breakdown,
but are interested in knowing the specific geographical and sector origin of 
the value added in exports, we van use the command `make_exvadir`: 

```{r make_exvadir, results='hide'}
exvadir <- make_exvadir(wio, exporter = "ESP", va_type = "FC", 
                        flow_type = "EXGR")
```

Please note that the `exvadir` object that we have obtained is different from
the `exvadec` object, in the sense that *'exporters'* in an `exvadir` object
are the different countries and sectors of origin of the value added included
in the exports of the country specified with `make_exvadir()`(in this case, 
Spain). We can better understand this by typing `summary(exvadir)`:

```{r summary exvadir}
summary(exvadir)
```

### Get specific data with custom groups

We can use `get_data()` to summarize the foreign content of Spanish exports,
with a breakdown between EU and Non-EU origin (specifying a few countries)
and also distinguishing between goods (with utilities) and
services. We can also break down the destination of those exports between
EU and non-EU: 

```{r get_data_power}
get_data(exvadir, exporter = c("WLD", "EU27", "FRA",
                               "NONEU27", "USA"),
         sector = c("TOTAL", "GOODSWU", "SRVWC"),
         importer = c("WLD", "EU27", "NONEU27"))
```

On the other hand, the flexibility of `get_data()` allows for the creation of
custom-made groups of countries and/or sectors.

For instance, let's create a group of countries called `LATAM`, with Spain
and Mexico. We would just have to define the variable
in the current environment.

```{r define_aukus}
LATAM <- c("ESP", "MEX")
```

And now we can use it as a normal variable, just introducing it as `"LATAM"` (do 
not forget the double quotes). We will use the `wwz` decomposition and
extract the domestic value added in intermediates (`DVA_INT`) from `LATAM` to
`USA`. Mind that, if we use custom groups, we need to select the 
option `custom = TRUE` in `get_data()`.

```{r get_data_vax_latam}
get_data(exvawwz, "DVA_INT", exporter = "LATAM", 
         sector = c("TOTAL", "MANUF", "SRVWC"), 
         importer = "USA", custom = TRUE)
```

Let us now see an  example of the exception marker `"x"`, that allows to define
exceptions for countries and for sectors. We can, for instance, calculate
the NAFTA exports, both intra-regional and extra-regional, employing just two
sectors: non-services and services.

```{r get_data_vax_latam_intrarreg}
get_data(exvawwz, "EXGR", exporter = "NAFTA", 
         sector = c("TOTAL", "TOTALxSRVWC", "SRVWC"), 
         importer = c("WLD", "NAFTA", "WLDxNAFTA"), custom = TRUE)
```

### Other useful commands

The flexibility of the commands `make_exvadir()` and `get_data()` allows for
the creation of several ready-made commands in **exvatools**. One is
`get_va_exgr()`, to get a detailed sector and geographical origin and 
destination of value added. For instance, if we want the value added in
services of the US incorporated in the Spanish exports of 
manufactures, we just have to type:

```{r get_va_exgr}
get_va_exgr(wio,geo_orig = "USA", sec_orig = "SRVWC",
            geo_export = "ESP", sec_export = "MANUF")
```

Sometimes we are not only interested in the origin, but also in the country
of final absorption. For that we have `get_va_exgry()`. For instance, if we 
want to know what part of the US value added incorporated in China's exports
of manufactures end up absorbed back in the US, we can type:

```{r get_va_exgry}
get_va_exgry(wio, geo_orig = "USA", geo_export = "CHN",
             sec_export = "MANUF", geo_fd = "USA")
```

At the beginning we manually calculated the value added induced by final 
demand. There is also a specific command for that in **exvatools** called
`get_va_fd()`. This allows, for instance, the calculation of the Chinese total
value added (or GDP) induced by US final demand for manufactures:

```{r get_va_fd}
get_va_fd(wio, geo_orig = "CHN", sec_orig = "TOTAL",
          geo_fd = "USA", sec_fd = "MANUF")
```

### TiVA useful indicators

Finally, if we want to get a list of common trade indicators (exports, imports,
value added, production) similar to those of the TiVA database, we could just
use `make_exvadec()` with the method `"oecd"` and `output = "tiva"`.

```{r make_exvadec_oecd_tiva, results='hide'}
exvativa <- make_exvadec(wio, exporter = "all", method = "oecd",
                         output = "tiva", quiet = TRUE)
```

And then get the decomposition for Spain:
```{r tiva_esp}
get_exvadec_bkdown(exvativa, exporter = "ESP")
```

## Available sectors and countries in each input-output edition

To check the information about sectors, it suffices to print `info_sec()`:
```{r info_sec}
info_sec("iciotest")
```

To check the information about available countries, the command is `info_geo()`:
```{r info_geo}
info_geo("iciotest")
```

This commands do not require to have a `wio` in the environment, so we can just
check what sectors are available in the OECD's ICIO tables, 2021 edition.
```{r info_sec_2021}
info_sec("icio2021")
```

Additionally, the commands `get_geo_codes()` and `get_sec_codes` provide details
about the components of the different groups. These commands are also directly
applicable for any available input-output table. For instance, for `"wiod2016"`
we would have the following components of EU27:
```{r get_geo_codes_wiod2016}
get_geo_codes("EU27", wiotype = "wiod2016")
```
And for `"icio2021"` we have the following components of the business services
sector:
```{r get_sec_codes_icio2021}
get_sec_codes("BIZSV", wiotype = "icio2021")
```
